\documentclass[11pt,a4paper]{article}
\usepackage[margin=25mm]{geometry}
\usepackage{amsmath,amssymb,bm}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{microtype}

\title{\textbf{Adaptive Controller: Model and Stability Proof}\\
\large Adaptive aerial-refueling station keeping}
\author{}
\date{}

\begin{document}
\maketitle

\section{Reduced Relative Model}

The tanker-relative frame is fixed at the desired refueling point. The
controlled point is the receiver-side refueling point. The state is the
relative position $\bm r$ and relative velocity $\bm v$ in the
longitudinal-vertical plane:
\[
\bm r =
\begin{bmatrix}
x \\ z
\end{bmatrix},
\qquad
\bm v = \dot{\bm r} =
\begin{bmatrix}
v_x \\ v_z
\end{bmatrix},
\qquad
\bm u =
\begin{bmatrix}
u_T \\ u_L
\end{bmatrix}.
\]

After subtracting the nominal trimmed forces, the perturbation model is
\[
\dot{\bm r}=\bm v,
\qquad
m(t)\dot{\bm v}
=
\bm u-C\bm v-\bigl(m(t)-m_0\bigr)g\bm \xi_z,
\]
where
\[
C=\operatorname{diag}(c_x,c_z),
\qquad
\bm \xi_z=
\begin{bmatrix}
0 \\ 1
\end{bmatrix}.
\]

The unknown scalar parameter used by the adaptive controller is the inverse
mass:
\[
\theta(t)=\frac{1}{m(t)}.
\]
Then the acceleration equation can be written in the adaptive-control form
\[
\dot{\bm v}
=
\theta\bm\phi(\bm s,\bm u)-g\bm\xi_z,
\qquad
\bm\phi(\bm s,\bm u)
=
\bm u-C\bm v+m_0g\bm\xi_z.
\]

\section{Error Variable and Control Law}

Use the filtered tracking error
\[
\bm e=\bm v+\Lambda\bm r,
\qquad
\Lambda=\operatorname{diag}(\lambda_x,\lambda_z)>0.
\]
If $\bm e\to0$, then
\[
\dot{\bm r}=-\Lambda\bm r,
\qquad
\bm r(t)\to0.
\]

The derivative of the filtered error is
\[
\dot{\bm e}
=
\dot{\bm v}+\Lambda\bm v
=
\theta\bm\phi(\bm s,\bm u)-g\bm\xi_z+\Lambda\bm v.
\]

For known $\theta$, the target dynamics are $\dot{\bm e}=-K\bm e$.
Replacing $\theta$ by its estimate gives the certainty-equivalent controller
\[
\bm u
=
C\bm v-m_0g\bm\xi_z
+
\frac{1}{\hat\theta}
\left(
-K\bm e-\Lambda\bm v+g\bm\xi_z
\right),
\]
where
\[
K=\operatorname{diag}(k_x,k_z)>0.
\]

The estimate is projected to a physically meaningful inverse-mass interval:
\[
\hat\theta\in[\theta_{\min},\theta_{\max}],
\qquad
\hat m=\frac{1}{\hat\theta}.
\]

\section{Constant-Mass Lyapunov Proof}

First assume that the mass is unknown but constant, so $\theta$ is constant.
Define the parameter error
\[
\tilde\theta=\theta-\hat\theta.
\]
Substituting the adaptive control law into the plant gives
\[
\dot{\bm e}
=
-K\bm e+\tilde\theta\bm\phi(\bm s,\bm u).
\]

Choose the augmented Lyapunov function
\[
L(\bm e,\tilde\theta)
=
\frac{1}{2}\bm e^T\bm e
+
\frac{1}{2\gamma}\tilde\theta^2,
\qquad
\gamma>0.
\]
It is positive definite:
\[
L(\bm e,\tilde\theta)>0
\quad\text{for}\quad
(\bm e,\tilde\theta)\ne(0,0),
\qquad
L(0,0)=0.
\]

Since $\theta$ is constant,
$\dot{\tilde\theta}=-\dot{\hat\theta}$. Therefore,
\[
\dot L
=
\bm e^T(-K\bm e+\tilde\theta\bm\phi)
-
\frac{1}{\gamma}\tilde\theta\dot{\hat\theta},
\]
or equivalently
\[
\dot L
=
-\bm e^TK\bm e
+
\tilde\theta
\left(
\bm e^T\bm\phi
-
\frac{1}{\gamma}\dot{\hat\theta}
\right).
\]

Choose the adaptation law
\[
\dot{\hat\theta}
=
\gamma\bm e^T\bm\phi(\bm s,\bm u).
\]
Then the parameter-error cross term cancels:
\[
\dot L
=
-\bm e^TK\bm e
\le 0.
\]

Thus $L$ is nonincreasing, $\bm e$ and $\tilde\theta$ are bounded, and
$\bm e$ is square integrable. Under the standard boundedness assumptions used
in direct adaptive control, Barbalat's lemma implies
\[
\bm e(t)\to0.
\]
Since
\[
\dot{\bm r}
=
-\Lambda\bm r+\bm e
\]
and $\Lambda$ is positive definite, the relative position also converges:
\[
\bm r(t)\to0.
\]

\section{Slowly Time-Varying Mass}

During refueling, the mass is not exactly constant. The inverse mass
$\theta(t)=1/m(t)$ slowly varies because the fuel flow is bounded. In this
case
\[
\dot{\tilde\theta}
=
\dot\theta-\dot{\hat\theta},
\]
and the Lyapunov derivative becomes
\[
\dot L
=
-\bm e^TK\bm e
+
\frac{1}{\gamma}\tilde\theta\dot\theta.
\]

The first term is dissipative. The second term is a bounded disturbance term
caused by fuel transfer. Therefore the result is practical stability, or
ultimate boundedness: the tracking error remains bounded and small when the
mass changes slowly relative to the closed-loop response.

\section{Discrete Implementation}

The simulation uses the projected update
\[
\hat\theta_{k+1}
=
\operatorname{proj}_{[\theta_{\min},\theta_{\max}]}
\left(
\hat\theta_k
+
\Delta t\,\gamma\bm e_k^T\bm\phi(\bm s_k,\bm u_k)
\right).
\]
The reported adaptive mass estimate and saturated command are
\[
\hat m_k=\frac{1}{\hat\theta_k},
\qquad
\bm u_k
=
\operatorname{sat}_{[-u_{\max},u_{\max}]}
\left\{
\bm u(\bm s_k,\hat\theta_k)
\right\}.
\]

The saturation is used in simulation to keep the command physically
meaningful; the analytical proof above corresponds to the unsaturated
adaptive law.

\end{document}
